Mechanization of Separation in Generic Extensions

### 2019-01-09 LICS/conclusion_m.tex
Paulson tambien reflexiona
Neeman sobre reflexión. No me convence dónde lo puse

The Replacement Axiom, however, requires some more work to be
done. Both
in Kunen and in other presentations (e.g. \cite{neeman-course}) it
requires a relativized version (i.e., showing that it holds for $M$)
of the \emph{Reflection Principle}. In order to state this
meta-theoretic result by Montague, recall that an equivalent
formulation of the Foundation Axiom states that the universe of sets
can be decomposed in a transfinite hierarchy of sets:
\begin{theorem}
  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
  $\al$. Then each $V_\al$ is a set and 
  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
\end{theorem}
\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
  unboundedly many $\al$ such that $V_\al\models \Phi$.''
\end{theorem}
%% This result has also other applications; in particular, it dispenses
%% the need of ctms of the whole of $\ZF$ for consistency proofs (by
%% repeating  the proof of Lemma~\ref{lem:wf-model-implies-ctm}, now
%% starting with a  $V_\al$ satisfying $\Phi$).

% This statement is straightforwardly equivalent to that with $\Phi$
% consisting of a single formula. 
It is obvious that we can take the conjunction of $\Phi$ and state
Theorem~\ref{th:reflection-principle} for a single formula, say $\phi$.
The schematic nature of this result
hints at a proof by induction on formulas, and hence it must be shown
internally.

>>>>

The Replacement Axiom, however, requires some more work to be
done. In Kunen it
requires a relativized version (i.e., showing that it holds for $M$)
of the \emph{Reflection Principle}. In order to state this
meta-theoretic result by Montague, recall that an equivalent
formulation of the Foundation Axiom states that the universe of sets
can be decomposed in a transfinite, cumulative hierarchy of sets:
\begin{theorem}
  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
  $\al$. Then each $V_\al$ is a set and 
  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
\end{theorem}
\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
  unboundedly many $\al$ such that $V_\al\models \Phi$.''
\end{theorem}
%% This result has also other applications; in particular, it dispenses
%% the need of ctms of the whole of $\ZF$ for consistency proofs (by
%% repeating  the proof of Lemma~\ref{lem:wf-model-implies-ctm}, now
%% starting with a  $V_\al$ satisfying $\Phi$).

% This statement is straightforwardly equivalent to that with $\Phi$
% consisting of a single formula. 
It is obvious that we can take the conjunction of $\Phi$ and state
Theorem~\ref{th:reflection-principle} for a single formula, say $\phi$.
The schematic nature of this result
hints at a proof by induction on formulas, and hence it must be shown
internally. It is to be noted that Paulson
\cite{DBLP:conf/cade/Paulson02} also formalized the
Reflection principle in Isabelle/ZF, but it is not clear if the
relativized version follows directly from it. (It may be possible to
sidestep Reflection, since in Neeman \cite{neeman-course}, only the
relativization of the cumulative hierarchy is needed; nevertheless, it
is a nontrivial task.)
